---
title: "assignment 3- group work"
format: html
editor: visual
---

### Methodology

This analysis uses the **Bike Sharing Dataset** from *UCI Machine Learning Repository*, which provides hourly rental data from 2011 to 2012 in Washington D.C. The dataset includes over 17,000 observations, with no misisng values. The variables capturing weather conditions, temporal attributes, and user counts. Our objective is to explore how environmental and time-based features influence rental demand.

We categorized the variables into three main areas of interest. First, the **weather features** include temperature (`temp`), perceived temperature (`atemp`), humidity (`hum`), and windspeed (`windspeed`). These values are scaled between 0 and 1 based on predefined minimum and maximum values provided by the dataset authors.

Secondly, **temporal variables** include hour (`hr`), day of the week (`weekday`), and binary indicators for holidays and working days (`holiday`,`workingday`). Season (`season`) and month (`mnth`) are provided as categorical values. These features allow us to explore rental demand across time, including difference between weekdays and weekends, and variations by time of day.

Third, we analyzed the **user count features**, including total rentals (`cnt`), as well as `casual` and `registered` users. This breakdown helps identify usage patterns by user type, including subscription behavior and potential tourist or leisure activity.

The report was conducted using **R**, with `tidyverse` package for data wrangling and `ggplot2` for visualization. **Quarto** was used for reproducible report generation, and **Git** for collaborative version control across team branches.

@fig-season-users shows the average total user count by season, revealing strong seasonal effects. @tbl-user-type-by-season compares average casual and registered users, showing that registered users dominate usage but casual users increase during warmer months. Additional findings are presented in the *Results section*.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#| label: fig-season-users
#| fig-cap: "Average Total Users by Season"

library(tidyverse)

# Read the dataset (assuming you want to use the daily data for seasonal trends)
day_data <- read.csv("data/day.csv")

# Plot: Average total users by season
day_data %>%
  group_by(season) %>%
  summarise(avg_total = mean(cnt)) %>%
  ggplot(aes(x = factor(season), y = avg_total)) +
  geom_col(fill = "lightblue") +
  labs(
   title = "Average Total Users by Season",
    x = "Season (1=Winter, 2=Spring, 3=Summer, 4=Fall)",
    y = "Average Users"
  )

```

```{r,echo =FALSE}
#| label: tbl-user-type-by-season
#| tbl-cap: "Average Casual and Registered Users by Season"

day_data %>%
  group_by(season) %>%
  summarise(
    avg_casual = round(mean(casual), 1),
    avg_registered = round(mean(registered), 1)
  ) %>%
  knitr::kable()

```
