---
title: "From Data to Action: Leveraging Bike-Sharing Patterns to Enhance Sustainable Mobility"
author:
  - Izzul Fattah Aji Pratama
  - Thuy Nhu Nguyen
  - Bhuvana Chandrasekar
format:
  revealjs:
<<<<<<< HEAD
    theme: solarized        
    slide-number: true   
    toc: false
=======
    theme: solarized        # or try: default, serif, solarized, night
    slide-number: true   # shows current slide number
    toc: false           # no table of contents slide
>>>>>>> 8edc42eb610fd9af04343f3e0fbde2b543584e22
editor: visual
---

```{r setup, include=FALSE}
library(tidyverse)
day_data <- read.csv("data/day.csv")
hour_data <- read.csv("data/hour.csv")
```

## Title Slide

**From Data to Action: Leveraging Bike-Sharing Patterns**\
*ETC5513 – Week 12 Presentation*\
**Team:**\
- Izzul Fattah Aji Pratama\
- Thuy Nhu Nguyen\
- Bhuvana Chandrasekar

## Introduction

-   Urban transport is shifting toward sustainable alternatives\
-   Bike sharing promotes health, efficiency, and lower emissions\
-   Our goal: analyze rental trends and make recommendations

## Dataset Overview

-   **Source**: [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset)\
-   **Timeframe**: 2011–2012\
-   **Data**: `day.csv` and `hour.csv` — weather, time, users

## Methodology

-   Tools: R, tidyverse, ggplot2, Quarto, Git\
-   Explored trends by weather, season, and time\
-   Collaboration via GitHub with reproducible reports

## Weather Effects

-   Rentals increase with temperature and decrease with humidity\

-   Wind speed has a slight negative effect\

-   Visualization: *weather vs rentals*

    ```{r, warning=FALSE, message=FALSE, fig.width=6, fig.height=3.8}
    day_data %>%
      pivot_longer(cols = c(temp, atemp, hum, windspeed),
                   names_to = "feature", values_to = "value") %>%
      ggplot(aes(x = value, y = cnt)) +
      geom_point(alpha = 0.3, color = "#3182bd") +
      geom_smooth(method = "lm", se = FALSE, color = "darkred") +
      facet_wrap(~feature, scales = "free") +
      labs(title = "Weather Features vs. Total Rentals", x = "Value", y = "Total Rentals") +
      theme_bw()

    ```

## Hourly Trends

-   Peak rentals during commute hours: 8 AM and 5–6 PM\

-   Weekends show midday increases, more variability

    ```{r}
    hour_data %>%
      mutate(weektype = ifelse(weekday %in% c(0, 6), "Weekend", "Weekday")) %>%
      pivot_longer(cols = c(casual, registered), names_to = "user_type", values_to = "count") %>%
      group_by(hr, user_type, weektype) %>%
      summarise(avg_count = mean(count), .groups = "drop") %>%
      ggplot(aes(x = hr, y = avg_count, fill = user_type)) +
      geom_col(position = "stack") +
      facet_wrap(~weektype) +
      labs(
        title = "Hourly Rentals by User Type on Weekdays vs Weekends",
        x = "Hour of Day",
        y = "Average Rentals",
        fill = "User Type"
      ) +
      theme_bw()

    ```

## Seasonal & User Types

-   Casual users increase in summer\

-   Registered users dominate during winter\

    ```{r}
    day_data %>%
      pivot_longer(cols = c(casual, registered), names_to = "user_type", values_to = "count") %>%
      group_by(season, user_type) %>%
      summarise(avg_count = mean(count), .groups = "drop") %>%
      ggplot(aes(x = factor(season), y = avg_count, fill = user_type)) +
      geom_col(position = "dodge") +
      labs(title = "Average Users by Season", x = "Season (1 = Winter, 2 = Spring, 3 = Summer, 4 = Fall)", y = "Average Rentals", fill = "User Type") +
      theme_bw()

    ```

## Key Insights

| Feature            | Effect on Rentals                        |
|--------------------|------------------------------------------|
| Temperature        | Strong positive impact                   |
| Feels-like Temp    | Similar to temperature                   |
| Humidity           | Mild negative impact                     |
| Windspeed          | Weak correlation                         |
| Hour of Day        | Peaks during morning and evening commute |
| Weekday vs Weekend | Higher and consistent on weekdays        |
| Time Progression   | General upward trend year over year      |

## Recommendations

-   Add bikes in tourist zones during summer\
-   Prioritize commuters with reliable service\
-   Plan fleet and staff using hourly demand

## Thank You

Questions? We're happy to discuss our findings. Cheers!
