---
title: "assignment 3- group work"
format: html
editor: visual
---

### Methodology:

This analysis uses the **Bike Sharing Dataset** from the UCI Machine Learning Repository, containing two files: `day.csv`(daily rental data) and `hour.csv` (hourly rental data) from 2011 to 2012 in Washington D.C.

The datasets are complete, with no missing values. They contain weather metrics, timestamps, and detailed user counts. We categorized the variables into three analytical areas:

1\. **Environmental conditions**: Continuous variables include temperature (`temp`), perceived temperature (`atemp`), humidity (`hum`), and windspeed (`windspeed`) — all normalized. These were sourced from \`hour.csv\` and used to assess how weather influences demand.

2\. **Temporal factors**: These include hour of day (`hr`), day of week (`weekday`), holidays, and working days (`workingday`). `hour.csv`was used to examine hourly rental patterns and compare weekdays to weekends. Seasonal trends were analyzed using `season` (1 = Winter to 4 = Fall) and `mnth` from `day.csv`.

3\. **User behavior**: Total rentals (`cnt`) are split into `casual`(non-registered) and `registered` (subscribers). `day.csv` was used to study seasonal variation in user types, while `hour.csv` was used for finer temporal granularity.

All analysis was performed in **R**, using `tidyverse` for data wrangling and `ggplot2` for visualization. The report was built in **Quarto** for reproducibility, and **Git** was used for structured collaboration via separate branches.

@fig-season-users shows average total users by season using `day.csv`, capturing clear seasonal demand shifts.

@tbl-hourly-weektype uses `hour.csv` to show average hourly rentals on weekdays vs. weekends, demonstrating time-of-day behavioral differences. Further analysis appears in the Results section.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#| label: fig-season-users
#| fig-cap: "Average Total Users by Season"
library(tidyverse)

day_data <- read.csv("data/day.csv")

day_data %>%
  group_by(season) %>%
  summarise(avg_total = mean(cnt)) %>%
  ggplot(aes(x = factor(season), y = avg_total)) +
  geom_col(fill = "lightblue") +
  labs(
    title = "Average Total Users by Season",
    x = "Season (1=Winter, ..., 4=Fall)",
    y = "Average Users"
  )

```

```{r,echo =FALSE}
#| label: tbl-hourly-weektype
#| tbl-cap: "Average Hourly Rentals on Weekdays vs Weekends"

hour_data <- read.csv("data/hour.csv")

hour_data %>%
  mutate(weektype = ifelse(weekday %in% c(0, 6), "Weekend", "Weekday")) %>%
  group_by(hr, weektype) %>%
  summarise(avg_count = round(mean(cnt), 1), .groups = "drop") %>%
  pivot_wider(names_from = weektype, values_from = avg_count) %>%
  knitr::kable()


```

### Results:

#### 1. Seasonal Trends and User Behavior

To examine how bike usage shifts across seasons, we analyzed total rentals (`cnt`) and user type composition. @fig-user-by-season hows that demand peaks in summer and fall, while winter records the lowest activity. @tbl-user-share-by-season breaks down usage by user type, revealing that registered users dominate in all seasons but especially in winter, suggesting habitual commuting behavior. In contrast, casual user activity more than doubles in summer, indicating recreational or tourist usage. These seasonal shifts suggest that demand is driven by both weather and user intent, with commuters being more consistent and casual users more season-sensitive.

```{r,echo=FALSE}
#| label: fig-user-by-season
#| fig-cap: "Average Casual and Registered Users by Season"

day_data %>%
  pivot_longer(cols = c(casual, registered), names_to = "user_type", values_to = "count") %>%
  group_by(season, user_type) %>%
  summarise(avg_count = mean(count), .groups = "drop") %>%
  ggplot(aes(x = factor(season), y = avg_count, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(
    x = "Season (1=Winter, ..., 4=Fall)",
    y = "Average Rentals",
    fill = "User Type"
  )



```

```{r, echo=FALSE}
#| label: tbl-user-share-by-season
#| tbl-cap: "Percentage of Casual and Registered Users by Season"


day_data %>%
  group_by(season) %>%
  summarise(
    casual_total = sum(casual),
    registered_total = sum(registered),
    total = sum(cnt)
  ) %>%
  mutate(
    pct_casual = round(100 * casual_total / total, 1),
    pct_registered = round(100 * registered_total / total, 1)
  ) %>%
  select(season, pct_casual, pct_registered) %>%
  knitr::kable()

```

### Discussion, Conclusion and Recommendations:

This report set out to explore how environmental and temporal factors influence bike rental demand in Washington D.C., using the UCI Bike Sharing Dataset. We investigated rental patterns by season, weather, temperature, hour of day, and user type (casual vs. registered), with the goal of uncovering meaningful trends to inform decision-making.

Our results showed strong seasonal and hourly trends. Rental demand peaked in summer and fall, with significantly higher casual user activity in warmer months, suggesting weather-sensitive, possibly tourist-driven behavior. Registered users, in contrast, showed consistent usage throughout the year and across weekdays, indicating habitual commuting. Additionally, rental volumes varied by hour, with predictable peaks during morning and evening commute times on weekdays, and a more gradual midday rise on weekends.

These findings suggest that registered users represent a loyal, time-consistent base, while casual users drive demand spikes that are highly sensitive to weather and season. This distinction is critical for bike-sharing operators and city planners when allocating resources or tailoring marketing strategies.

We recommend:

-   Increasing bike availability and docking stations near recreational zones in summer

-   Prioritizing maintenance and reliability of service for commuters year-round

-   Using hourly and seasonal trends to optimize staffing, rebalancing logistics, and promotional campaigns

This analysis demonstrates how open data and reproducible tools can generate insights for urban mobility planning, infrastructure investment, and user engagement strategies — and the same framework can be applied to more recent datasets from other cities or systems.
