---
title: "assignment 3- group work"
format: html
editor: visual
---
```{r setup, include=FALSE}
# Load libraries and read datasets
library(tidyverse)
library(lubridate)
library(ggplot2)
```
```{r}}
day_data <- read_csv("data/bike+sharing+dataset/day.csv") %>%
  mutate(date = ymd(dteday),
         day_type = if_else(weekday %in% c(0, 6), "Weekend", "Weekday"))

hour_data <- read_csv("data/bike+sharing+dataset/hour.csv") %>%
  mutate(hr = as.factor(hr))
```

## Introduction

This report explores features affecting bike rentals:

-   **Weather conditions**: `temp`, `atemp`, `hum`, `windspeed`

-   **Time-based variables**: `hour`, `weekday`, `workingday`, `holiday`

------------------------------------------------------------------------

## Data Structure

```{r}
head(day_data)
head(hour_data)
```


## Hourly Rental Patterns

```{r}
hour_data %>%
  group_by(hr) %>%
  summarise(avg_rentals = mean(cnt)) %>%
  ggplot(aes(x = hr, y = avg_rentals)) +
  geom_col(fill = "skyblue") +
  labs(title = "Average Hourly Bike Rentals", x = "Hour", y = "Average Rentals") +
  theme_minimal()
```

The bar chart titled **"Average Hourly Bike Rentals"** presents the mean number of bike rentals across each hour of the day, offering valuable insight into daily usage patterns.

------------------------------------------------------------------------

### **Key Insights from the Chart:**

#### 1. **Two Prominent Peaks:**

-   **Morning Peak (\~8 AM):**

    -   Clear spike in average rentals around 8 AM.
    -   Likely corresponds to **commuting hours** for work or school.

-   **Evening Peak (\~5–6 PM):**

    -   Highest peak occurs around 5–6 PM.
    -   Reflects **commute back home** or post-work travel.

#### 2. **Midday Stability:**

-   Between **11 AM and 3 PM**, rentals are steady but lower than peak times.
-   May include **lunch breaks**, casual or leisure usage.

#### 3. **Low Usage at Night:**

-   From **midnight to 5 AM**, rentals are minimal.
-   Expected due to **lower demand and limited activity** during nighttime.

#### 4. **Gradual Build-up:**

-   Rentals start increasing sharply from around **6 AM**, indicating the **start of the day’s activity**.

------------------------------------------------------------------------

### **Use:**

#### **Operational Efficiency:**

-   **Rebalancing Bikes**: Optimize bike placement and availability ahead of peak hours.
-   **Maintenance Scheduling**: Perform maintenance during low-usage hours (e.g., 2–4 AM).

#### **User Behavior Understanding:**

-   Confirms that the system is heavily used by **commuters**, emphasizing the importance of **rush hour planning**.

#### **Infrastructure Planning:**

-   City planners or bike-sharing companies can plan **dedicated bike lanes or docking stations** where demand spikes.

------------------------------------------------------------------------

### Summary:

This hourly analysis highlights a **classic bimodal usage pattern** centered around commuting hours. Understanding this pattern helps improve user experience, resource allocation, and urban transport integration.

------------------------------------------------------------------------

## Weekday vs Weekend

This graph compares rental trends on weekdays vs. weekends across the full time period.

```{r}
# Weekday vs Weekend
 ggplot(day_data, aes(x = date, y = cnt, color = day_type)) +
  geom_line(alpha = 0.7) +
  geom_smooth(se = FALSE, method = "loess", span = 0.2) +
  scale_color_manual(values = c("Weekday" = "#1b9e77", "Weekend" = "#d95f02")) +
  labs(title = "Weekday vs Weekend Rentals", x = "Date", y = "Rentals", color = "Day Type")
```

The graph titled **"Weekday vs Weekend Rentals"** provides a time-series comparison of bike rental trends across weekdays and weekends. Let’s explore the key insights and their significance:

------------------------------------------------------------------------

### **Insights from the Graph:**

#### 1. **Overall Trend:**

-   Both weekday and weekend rentals follow a **seasonal pattern**, with higher usage during warmer months (peaking mid-year) and lower usage during colder months (start and end of the year).
-   The trends repeat over two years, suggesting strong seasonality in bike rentals.

#### 2. **Weekday Rentals (Green Line):**

-   Generally **higher and more consistent** than weekend rentals.
-   The green smoothed line shows stable high points during peak months (summer), likely due to daily commuting.

#### 3. **Weekend Rentals (Orange Line):**

-   More **variable and spiky**, possibly influenced by weather, events, or leisure activities.
-   Although lower than weekday rentals on average, weekends still show strong peaks during favorable months.

### Summary:

This graph clearly shows that **bike rental patterns differ significantly between weekdays and weekends**, with **seasonality and consistency** being key differences. It helps identify **when and how users engage** with the service, which is crucial for data-driven decision making in urban mobility planning or bike-sharing operations.

```{r}

# Plot: Date vs Rental Count
ggplot(day_data, aes(x = dteday, y = cnt)) +
  geom_line(color = "#1f77b4", size = 1) +
  labs(
    title = "Daily Bike Rentals Over Time",
    x = "Date",
    y = "Total Rentals"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

1.  **Clear Upward Trend in First Year**:

Rentals steadily increase from early 2011 through mid-2011. Indicates growing adoption of the bike-sharing service, possibly due to awareness or seasonal effects.

2.  **Strong Seasonal Patterns**:

Peaks are consistently seen during summer months (June–August). Dips in usage appear in winter months, likely due to colder weather and reduced ridership.

3.  **Sharp Spikes and Drops**:

Several days show sudden drops (close to 0) and spikes: Likely due to missing data, system outages, or extreme weather. Worth investigating further using weather data or logs.

4.  **Higher Plateau in Second Year**:

In 2012, daily rentals are consistently higher compared to 2011. This suggests sustained user base growth and possibly expansion of infrastructure.

```{r}
# Summer 2011 Focus
 day_data %>%
  filter(date >= "2011-06-01", date <= "2011-08-31") %>%
  ggplot(aes(x = date, y = cnt)) +
  geom_line(color = "#31a354") +
  geom_smooth(se = FALSE, method = "loess", span = 0.4, color = "#e34a33") +
  labs(title = "Bike Rentals in Summer 2011", x = "Date", y = "Rentals")
```

What the Chart Shows: Green Line: Raw daily bike rental counts. Red LOESS Curve: Smoothed trend to highlight overall direction (captures nonlinear patterns).

### Key Observations:

**Relatively Stable Start (June)**:

Rentals fluctuate around 4800–5200. Early signs of stability as the program matured post-launch.

**Rising Trend Toward July**: Clear upward drift through mid-to-late June into July. Possibly reflects better weather, longer daylight hours, or increased adoption.

**Mid-July Peak**: Several days around 6000 rentals—among the highest in the summer. Could correspond to events or peak seasonal demand.

**August Dip**: Rentals decline in early-to-mid August. A few drops below 4000 rentals—possibly due to weather events, vacations, or operational issues.

**End-of-August Anomaly**: A sudden sharp drop near end of August (close to 0 rentals). This is likely a data gap, outage, or recording error.

### Use:

**Operational Planning**: Staffing and maintenance could be scaled with demand (peaking July).

**Event/Promotion Timing**: Best results likely mid-summer.

**Anomaly Handling**: Investigate sharp drops for data integrity.

------------------------------------------------------------------------

## Summary of Insights

| Feature            | Effect on Rentals                        |
|--------------------|------------------------------------------|
| Temperature        | Strong positive impact                   |
| Feels-like Temp    | Similar to temperature                   |
| Humidity           | Mild negative impact                     |
| Windspeed          | Weak correlation                         |
| Hour of Day        | Peaks during morning and evening commute |
| Weekday vs Weekend | Higher and consistent on weekdays        |
| Time Progression   | General upward trend year over year      |

------------------------------------------------------------------------
