---
title: "assignment 3- group work"
format: html
editor: visual
---

### Methodology:

This analysis uses the **Bike Sharing Dataset** from the UCI Machine Learning Repository, containing two files: `day.csv`(daily rental data) and `hour.csv` (hourly rental data) from 2011 to 2012 in Washington D.C.

The datasets are complete, with no missing values. They contain weather metrics, timestamps, and detailed user counts. We categorized the variables into three analytical areas:

1\. **Environmental conditions**: Continuous variables include temperature (`temp`), perceived temperature (`atemp`), humidity (`hum`), and windspeed (`windspeed`) — all normalized. These were sourced from \`hour.csv\` and used to assess how weather influences demand.

2\. **Temporal factors**: These include hour of day (`hr`), day of week (`weekday`), holidays, and working days (`workingday`). `hour.csv`was used to examine hourly rental patterns and compare weekdays to weekends. Seasonal trends were analyzed using `season` (1 = Winter to 4 = Fall) and `mnth` from `day.csv`.

3\. **User behavior**: Total rentals (`cnt`) are split into `casual`(non-registered) and `registered` (subscribers). `day.csv` was used to study seasonal variation in user types, while `hour.csv` was used for finer temporal granularity.

All analysis was performed in **R**, using `tidyverse` for data wrangling and `ggplot2` for visualization. The report was built in **Quarto** for reproducibility, and **Git** was used for structured collaboration via separate branches.

@fig-season-users shows average total users by season using `day.csv`, capturing clear seasonal demand shifts. @tbl-hourly-weektype uses `hour.csv` to show average hourly rentals on weekdays vs. weekends, demonstrating time-of-day behavioral differences. Further analysis appears in the Results section.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#| label: fig-season-users
#| fig-cap: "Average Total Users by Season (from day.csv)"
library(tidyverse)

day_data <- read.csv("data/day.csv")

day_data %>%
  group_by(season) %>%
  summarise(avg_total = mean(cnt)) %>%
  ggplot(aes(x = factor(season), y = avg_total)) +
  geom_col(fill = "lightblue") +
  labs(
    title = "Average Total Users by Season",
    x = "Season (1=Winter, ..., 4=Fall)",
    y = "Average Users"
  )

```

```{r,echo =FALSE}

#| label: tbl-hourly-weektype
#| tbl-cap: "Average Hourly Rentals on Weekdays vs Weekends (from hour.csv)"


hour_data <- read.csv("data/hour.csv")

hour_data %>%
  mutate(weektype = ifelse(weekday %in% c(0, 6), "Weekend", "Weekday")) %>%
  group_by(hr, weektype) %>%
  summarise(avg_count = round(mean(cnt), 1), .groups = "drop") %>%
  pivot_wider(names_from = weektype, values_from = avg_count) %>%
  knitr::kable()


```

### Results:

#### 1. Seasonal Trends and User Behavior

To examine how bike usage shifts across seasons, we analyzed total rentals (`cnt`) and user type composition. @fig-user-by-season hows that demand peaks in summer and fall, while winter records the lowest activity. @tbl-user-share-by-season breaks down usage by user type, revealing that registered users dominate in all seasons but especially in winter, suggesting habitual commuting behavior. In contrast, casual user activity more than doubles in summer, indicating recreational or tourist usage. These seasonal shifts suggest that demand is driven by both weather and user intent, with commuters being more consistent and casual users more season-sensitive.

```{r,echo=FALSE}
#| label: fig-user-by-season
#| fig-cap: "Average Casual and Registered Users by Season"

day_data %>%
  pivot_longer(cols = c(casual, registered), names_to = "user_type", values_to = "count") %>%
  group_by(season, user_type) %>%
  summarise(avg_count = mean(count), .groups = "drop") %>%
  ggplot(aes(x = factor(season), y = avg_count, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(
    x = "Season (1=Winter, ..., 4=Fall)",
    y = "Average Rentals",
    fill = "User Type"
  )



```

```{r, echo=FALSE}
#| label: tbl-user-share-by-season
#| tbl-cap: "Percentage of Casual and Registered Users by Season"


day_data %>%
  group_by(season) %>%
  summarise(
    casual_total = sum(casual),
    registered_total = sum(registered),
    total = sum(cnt)
  ) %>%
  mutate(
    pct_casual = round(100 * casual_total / total, 1),
    pct_registered = round(100 * registered_total / total, 1)
  ) %>%
  select(season, pct_casual, pct_registered) %>%
  knitr::kable()

```
